---
title: "B-Splines and Control Polygons"
author: "Peter E. DeWitt"
output:
 rmarkdown::html_vignette:
   toc: true
   number_sections: true
bibliography: references.bib
vignette: >
 %\VignetteIndexEntry{B-Splines and Control Polygons}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---

<style type="text/css">
window.MathJax = {
   tex: {
     tags: 'ams'
   }
};
</style>


```{r label = "setup", include = FALSE}
library(qwraps2)
options(qwraps2_markup = "markdown")
knitr::opts_chunk$set(collapse = TRUE)
```



```{r}
library(cpr)
packageVersion("cpr")
```


The term "spline" is likely derived from shipwright or draftsmen splines,
thin wood strips, held in place by weights, used to define curves.  These
splines naturally minimize strain energy and the use of additional weights at
strategic locations on the spline are needed to achieve specific curvatures.
Cubic B-splines are not dissimilar.

Splines are piecewise polynomial curves that are differentiable up to a
prescribed order. B-splines are based on a _basis_ of polynomial functions.

Definitions and notation for uni-variable and multi-variable B-splines and
the associated control polygons and control nets are presented in the
following.  Two very good references for splines are @deboor2001 and
@prautzsch2002 if you wish to dig into the details.

# Uni-variable B-splines and Control Polygons

## B-splines

A curve defined by $f\left(x\right)$ is a spline of order $k$ (degree $k
-1$), with knots $\xi_1, \xi_2, \ldots, \xi_m$ where $\xi_i \leq x_{i+1}$ and
$\xi_{i} < \xi_{i + k} \forall i$ if $f\left(x\right)$ is $k-r-1$ times
differentiable at any $r$-fold knot, and $f\left(x\right)$ is a polynomial of
order $\leq k$ over each interval $x \in \left[\xi_i, \xi_{i+1}\right]$ for
$i = 0, \ldots, m - 1.$

In particular, B-splines, are defined as an affine combination:

\begin{equation}
  f \left( x \right) =
  \sum_{j} \theta_j B_{j,k,\boldsymbol{\xi}}\left(x\right) =
  \boldsymbol{B}_{k, \boldsymbol{\xi}} \left(x\right) \boldsymbol{\theta}_{\boldsymbol{\xi}}
  \label{eq:f}
\end{equation}

where $B_{j,k,\boldsymbol{\xi}}\left(x\right)$ is the $j^{th}$ basis spline
function, $\boldsymbol{\xi}$ is a sequence of $l$ interior knots and a total
of $2k$ boundary knots, i.e., $\left\lvert \boldsymbol{\xi}
\right\rvert| = 2k + l.$ The value of the $k$ boundary knots is arbitrary,
but a common choice is to use $k$-fold knots on the boundary:

$$ \xi_1 = \xi_2 = \cdots = \xi_k < \xi_{k+1} \leq \cdots \leq \xi_{k + l} =
\xi_{k + l + 1} = \cdots \xi_{2k + l}.$$

Alternative boundary knots can be used so long as the sequence
$\boldsymbol{\xi}$ is non-decreasing.  More on the implications of $k$-fold
boundary knots follow in the next section.

The $j^{th}$ B-spline is defined as:

\begin{equation}
B_{j,k,\boldsymbol{\xi}}\left(x\right) =
 \omega_{j,k,\boldsymbol{\xi}} \left(x\right)
 B_{j,k-1,\boldsymbol{\xi}}\left(x\right) +
 \left(1 - \omega_{j+1,k,\boldsymbol{\xi}} \right)
 B_{j+1,k-1,\boldsymbol{\xi}}\left(x\right),
\end{equation}
where
\begin{equation}
B_{j,k,\boldsymbol{\xi}}\left(x\right) = 0 \quad \text{for} \quad x \notin
\left[\xi_j, \xi_{j+k} \right), \quad \quad
B_{j,1,\boldsymbol{\xi}}\left(x\right) = \begin{cases}
1 & x \in \left[\xi_j, \xi_{j+k}\right) \\ 0 & \text{otherwise},
\end{cases}
\end{equation}
and
\begin{equation}
\omega_{j,k,\boldsymbol{\xi}}\left(x\right) = \begin{cases}
0 & x \leq \xi_j \\ \frac{x-\xi_j}{\xi_{j+k-1} - \xi_{j}} & \xi_j < x < \xi_{j+k-1} \\
1 & \xi_{j+k-1} \leq x.
\end{cases}
\end{equation}

For a set of observations, $\boldsymbol{x} = x_1, \ldots, x_n,$ the basis
functions defined above generalize to a matrix:

\begin{equation}
  \boldsymbol{B}_{k, \boldsymbol{\xi}} \left(\boldsymbol{x}\right) =
  \begin{pmatrix}
    B_{1, k, \boldsymbol{\xi}} \left(x_1\right) &
    B_{2, k, \boldsymbol{\xi}} \left(x_1\right) &
    \ldots &
    B_{k + l, k, \boldsymbol{\xi}} \left(x_1\right) \\
    \vdots & \vdots & \ddots & \vdots \\
    B_{1, k, \boldsymbol{\xi}} \left(x_n\right) &
    B_{2, k, \boldsymbol{\xi}} \left(x_n\right) &
    \ldots &
    B_{k + l, k, \boldsymbol{\xi}} \left(x_n\right).
  \end{pmatrix}
\end{equation}

A couple of many important properties of the basis function is that for any value
of $x \in \left[\xi_1, \xi_{2k+l}\right]$

* $B_{j,k,\boldsymbol{xi}} \left(x\right) \in \left[0, 1\right],$
* $ B_{j,k,\boldsymbol{\xi}} \left(x) > 0 \quad \text{for} \quad x \in \left(\xi_{j}, \xi_{j+k}\right),$ and
* $\sum_j B_{j,k,\boldsymbol{xi}} \left(x\right) = 1.$

Within the cpr package we can generate a basis matrix thusly:

```{r label = "basis_matrix"}
x <- seq(0 + 1/5000, 6 - 1/5000, length.out = 5000)
bmat <- bsplines(x, iknots = c(1, 1.5, 2.3, 4, 4.5), bknots = c(0, 6))
bmat
```


Note: the default order for
`r backtick(bsplines) `
is 4, and the default for the boundary knots is the range of
`r backtick(x) %s% "." `


```{r label = "row sums to 1"}
all(bmat >= 0)
all(bmat <= 1)
all.equal(rowSums(bmat), rep(1, nrow(bmat)))
```


We can quickly view the plot of each of these spline functions as well.

```{r label = "plot bmat", fig.width = 7, fig.height = 4}
plot(bmat, show_xi = TRUE, show_x = TRUE)
```


## Spline Function Control Polygons

The spline $f\left(\boldsymbol{x}\right) = \boldsymbol{B}_{k,\boldsymbol{\xi}}\left(\boldsymbol{x}\right)$
is a convex sum of the coefficients $\boldsymbol{\theta}_{\boldsymbol{\xi}}.$
A meaningful geometric
relationship between $\boldsymbol{\xi}$ and
$\boldsymbol{\theta}_{\boldsymbol{\xi}}$ exist in the form of a
__control polygon__,
$CP_{k,\boldsymbol{\xi},\boldsymbol{\theta}_{\boldsymbol{\xi}}},$ a strong
convex hull for
$\boldsymbol{B}_{k,\boldsymbol{\xi}}\left(\boldsymbol{x}\right)
\boldsymbol{\theta}_{\boldsymbol{\xi}}.$

\begin{equation}
CP_{k,\boldsymbol{\xi},\boldsymbol{\theta}_{\xi}} =
\left\{ \left( \xi_{j}^{*}, \theta_{j,\boldsymbol{\xi}} \right) \right
\}_{j=1}^{\left\lvert\boldsymbol{\theta}_{\boldsymbol{\xi}}\right\rvert};
\quad \quad
\xi_{j}^{*} = \frac{1}{k-1} \sum_{i = 1}^{k-1} \xi_{j + i}.
\end{equation}

$CP_{k,\boldsymbol{\xi},\boldsymbol{\theta}_{\boldsymbol{\xi}}}$ is a
sequence of $\left\lvert \boldsymbol{\theta}_{\boldsymbol{\xi}} \right\rvert
= k + l$ control vertices.  The control polygon can be thought of as a
piecewise linear function approximating the spline function.  Changes in
convexity and other subtle characteristics of the spline function are
exaggerated by the control polygon.

For example, using the basis matrix defined above and the following
coefficients we can easily define a spline function and control polygon.

```{r label = "define theta"}
theta <- matrix(c(1, 0, 3.5, 4.2, 3.7, -0.5, -0.7, 2, 1.5), ncol = 1)
cp(bmat, theta)
```


Plotting the control polygon and the corresponding spline:


```{r label = "plot cp", fig.width = 7, fig.height = 4}
plot(cp(bmat, theta), show_spline = TRUE)
```


## Continuity and Knots

All the basis functions are, by definition, right-continuous.  The spline
function has at most $k$ continuity conditions.  That, is, for a value of $x$
such that $\xi_1 < x < \xi

The knots sequence is a representation of the desired smoothness of
$f\left(x\right).$  For any point $\xi_1 < x < \xi_{2k + l}} and $\notexists
i s.t. x = \xi_i,$ the function $f\left(x\right)$ has $k$ continuity
conditions, $(k -1)^{st}-$ through $0^{th}$-order differentiable.  At a knot,
there number of continuity conditions decreases for the number of knots at
that point.  Overall, the number of continuity conditions at a point $x$ plut
the number of knots at $x$ equals $k.$

For an example, we will use a $k = 4$ order spline (a cubic spline) to
perfectly fit a cubic function.  We'll define the function, first, and second
derivatives

```{r}
f <- function(x) {
  #(x + 2) * (x - 1) * (x - 3)
  x^3 - 2 * x^2 - 5 * x + 6
}

fprime <- function(x) { # first derivatives of f(x)
  3 * x^2 - 4 * x - 5
}

fdoubleprime <- function(x) { # second derivatives of f(x)
  6 * x - 4
}
```


We'll look at the function over

```{r}
bknots = c(-3, 5)

x     <- seq(-3 + 1/100, 5 - 1/100, length.out = 100)
bmat  <- bsplines(x, bknots = bknots)
theta <- matrix(coef(lm(f(x) ~ bmat + 0)), ncol = 1)

bmatD1 <- bsplineD(x, bknots = bknots, derivative = 1L)
bmatD2 <- bsplineD(x, bknots = bknots, derivative = 2L)

# check that the function f(x) is recovered
all.equal(f(x), as.numeric(bmat %*% theta))

# check that the first derivative is recovered
all.equal(fprime(x), as.numeric(bmatD1 %*% theta))

# check that the second derivative is recovered
all.equal(fdoubleprime(x), as.numeric(bmatD2 %*% theta))

# Plot the results
par(mfrow = c(1, 3))
plot(x, f(x), type = "l")
points(x, bmat %*% theta, col = 'blue')

plot(x, fprime(x), type = "l")
points(x, bmatD1 %*% theta, col = 'blue')

plot(x, fdoubleprime(x), type = "l")
points(x, bmatD2 %*% theta, col = 'blue')
```



# References
<div id="refs"></div>

# Session Info

```{r label = "sessioninfo"}
sessionInfo()
```

