---
title: "The Control Polygon Reduction Package"
author: "Peter DeWitt"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{The Control Polygon Reduction Package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(knitr)
opts_chunk$set(collapse = TRUE)
```

The cpr package provides several tools for finding parsimonious B-spline
regression models via control polygon reduction.  This vignette is an overview
of the tools provided by the package.

```{r}
library(cpr)
```

# CPR: Control Polygon Reduction
*Coming Soon*  A description of the CPR method, including the background on
B-splines and assessing knot influence will be provided soon.


# CNR: Control Net Reduction
*Coming Soon*  Control Net reduction is the natural extension of CPR from
uni-variable functions to multi-variable functions.  Details on this work will
be added to this vignette soon.

# B-Splines
Base R includes the splines package and the `splines::bs` call for building
B-splines.  The cpr package provides the `cpr::bsplines` call as there are
certain default behaviors and additional required meta-data storage requirements
needed for the CPR method that the `splines::bs` call does not provide.  In this
section we compare `splines::bs` and `cpr::bsplines` so that end users can
translate between the two functions.

```{r}
bmat <- bsplines(x = seq(0, 6, length = 500), iknots = c(1.0, 1.5, 2.3, 4.0, 4.5))
plot(bmat)
```

# Construction of Control Polygons

# Tensor-Products

## In General

## Of B-Splines Bases

# Construction of Control Nets


# Assessing the Relative Influence of a Knot.

# Running the CPR Model Selection Algorithm

## Plotting Diagnostics

## Plotting Curves

# Running the CNR Model Selection Algorithm

## Plotting Diagnostics

## Plotting Surfaces
